---
title: "User Manual: IgGeneUsage"
author: "SK"
date: "June 25, 2019"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{User Manual: IgGeneUsage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = FALSE, 
                      warning = FALSE, message = FALSE)
```



```{r}
require(knitr)
require(ggplot2)
require(ggforce)
require(gridExtra)
require(ggrepel)
require(rstan)
require(tidyr)
require(reshape2)
rstan_options(auto_write = TRUE)

source(file = "R/Util.R")
source(file = "R/Usage.R")
```





# Introduction
Decoding the properties of the immune repertoires is key in understanding the 
response of adaptive immunity during health challenges such as infection or 
disease. One important feature of all repertoires is their gene usage, i.e. the 
frequency at which the various germline genes have been rearranged to generate 
diverse B and T cell receptors (BCRs and TCRs).

If we discover a disruption in gene usage under a specific biological condition 
compared to a control, we can use this as hint about the potential link between 
that condition (disease, virus, etc.) and the immune mechanisms that for BCR/TCR 
generation and rearrangement, including subsequent selection pressures. 

The first step is, however, to detect differential gene usage between two 
biological conditions (e.g. healthy vs tumor). IgGeneUsage can help you do that.


# Usage
All you need to carry out your analysis with IgGeneUsage is a data.frame with 
the following 4 columns:

  1. sample_id: character identifier of the sample/repertoire (e.g. Patient-1)
  2. condition: character identifier of the condition to which each repertoire 
  belongs (e.g. healthy or tumor)
  3. gene_name: specific gene name (e.g. IGHV1-10 or family TRVB1)
  4. gene_usage_count: numeric (count) of rearrangements related to columns 1-3

Important remarks: 

  * the sum of the gene_usage_count (column 4) for a given sample (sample_id) 
  should be equal to the total number of rearrangements in that repertoire. 
  * the remaining parameters provided to IgGeneUsage help the MCMC sampler. The 
  default settings should suffice for most analyses.



# Case Study I
The IgGeneUsage package comes with built-in datasets of gene usage. In the first 
case study we will evaluate disruptions in the IGHV-segment usage of class 
switched (CS) memory B-cells between HCV+ (N = 22) and healthy individuals 
(N = 7)[^1].

[^1]: Tucci, Felicia A., et al. "Biased IGH VDJ gene repertoire and clonal 
expansions in B cells of chronically hepatitis C virus–infected individuals." 
Blood 131.5 (2018): 546-557.]


## Input data

```{r}
data("IGHV_HCV")
```


```{r}
kable(x = head(IGHV_HCV), row.names = FALSE)
```



## Data visualization
Lets look at the input data with ggplot2.

```{r, fig.height = 6, fig.wide = TRUE}
# we can compute the total number of rearrangements per sample
total.usage <- aggregate(gene_usage_count~sample_id+condition, 
                         FUN = sum, data = IGHV_HCV)
total.usage$total <- total.usage$gene_usage_count
total.usage$gene_usage_count <- NULL

# merge it with the original data
viz <- merge(x = IGHV_HCV, y = total.usage, 
                  by = c("sample_id", "condition"), 
                  all.x = TRUE)

# compute %
viz$gene_usage_pct <- viz$gene_usage_count/viz$total*100

# visualize
ggplot(data = viz)+
  geom_point(aes(x = gene_name, y = gene_usage_pct, 
                 fill = condition, shape = condition),
             position = position_dodge(width = .7), stroke = 0)+
  theme_bw(base_size = 12)+
  ylab(label = "Usage [%]")+
  xlab(label = '')+
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4))+
  scale_fill_manual(name = "Condition", values = c("orange", "#6666ff"))+
  scale_shape_manual(name = "Condition", values = c(21, 22))
```


## Analysis with IgGeneUsage
This is the main method of IgGeneUsage for differential gene usage.

```{r}
# M <- diffUsage(usage.data = IGHV_HCV,
#                mcmc.warmup = 500,
#                mcmc.steps = 1500,
#                mcmc.chains = 4,
#                mcmc.cores = 4,
#                hdi.level = 0.95,
#                adapt.delta = 0.95,
#                max.treedepth = 13,
#                dev.model = "src/stan_files/beta_binomial_model.stan")
# save(M, file = "R/dev/saved_temp/M_IGHV_HCV.RData")
M <- get(load(file = "R/dev/ighv_hcv_zibb_model.RData"))
```




## Model checking
Do not blindly trust the model $\rightarrow$ check it extensively. 

  * Checklist of successful MCMC sampling[^2]:
      * no divergences
      * no additional warnings from rstan
      * Rhat < 1.05 (see object glm.stats)
      * sufficiently high Neff
  * Checklist for valid model:
      * posterior predictive check reveal decent predictive accuracy on 
        already observed data (and also new data)
      * leave-one-out analysis with R-package loo (most k values should be low)

[^2]: https://mc-stan.org/misc/warnings.html

### MCMC sampling

  * divergences, tree-depth, energy
  
```{r}
rstan::check_hmc_diagnostics(M$glm)
```

  * Rhat and Neff

```{r, fig.height = 3, fig.wide = TRUE}
grid.arrange(rstan::stan_rhat(object = M$glm),
             rstan::stan_ess(object = M$glm),
             nrow = 1)
```


### Posterior predictive checks: repertoire-specific

  * IgGeneUsage performs automatic posterior predictive check (PPC).
  * In the following figure, we show the predictions (y-axis) of the model for 
  the usage of each gene (shown as point) in a given repertoire. The predictions 
  are compared against the observed data (x-axis).
  * If the point lies on the the diagnoal $\rightarrow$ accurate prediction.

```{r, fig.height = 12, fig.wide = TRUE}
ggplot(data = M$ppc)+
  facet_wrap(facets = ~sample_id, ncol = 5)+
  geom_abline(intercept = 0, slope = 1)+
  geom_errorbar(aes(x = raw, y = ppc.raw.mean, ymin = ppc.raw.L, 
                    ymax = ppc.raw.H), col = "darkgray")+
  geom_point(aes(x = raw, y = ppc.raw.mean, fill = condition), 
             shape = 21, size = 2)+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  scale_x_log10()+
  scale_y_log10()+
  xlab(label = "Observed usage [counts]")+
  ylab(label = "Predicted usage [counts]")
```




### Posterior predictive checks: overall differential usage

  * IgGeneUsage performs automatic posterior predictive check (PPC) at level of 
  genes as well, i.e. , predicts the overal (pooled) usage of a gene within each 
  condition.
  * Legend:
      * x-axis: mean usage of repertoires with given condition
      * y-axis: predicted usage (with 95% HDI)
  * If the point lies on the the diagnoal $\rightarrow$ accurate prediction.
  * In the previous plot, the usage data and the predictions consist counts. In 
  this plot, the usage is shown as % usage.

```{r, fig.height=5, fig.width = 6}
ggplot(data = M$group.ppc$group.ppc)+
  geom_abline(intercept = 0, slope = 1)+
  geom_errorbar(aes(x = raw.M, ymin = ppc.L, ymax = ppc.H), col = "darkgray")+
  geom_point(aes(x = raw.M, y = ppc.M, fill = condition), shape = 21)+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  xlab(label = "Observed usage [%]")+
  ylab(label = "Predicted usage [%]")
```



### Leave-one-out
  * with R-package loo
  
```{r}
loo::loo(x = loo::extract_log_lik(stanfit = M$glm))
```


## Results
This following figure shows the main results of IgGeneUsage $\rightarrow$ 
probability of differential gene usage ($\pi$) shown on the y-axis for different 
genes shown as points. The names of genes with $\pi > 0.95$ are shown.

```{r, fig.height=4, fig.wide = TRUE}
# format data
stats <- M$glm.summary
stats <- stats[order(abs(stats$effect_mean), decreasing = F), ]
stats$gene_fac <- factor(x = stats$gene_name, levels = stats$gene_name)

stats <- merge(x= stats, y = M$test.stats, by = "gene_name")

ggplot()+
  geom_point(data = stats[stats$pmax < 0.95, ], aes(y = pmax, x = ''),
             position = position_jitter(width = 0.25, height = 0))+
  geom_text(data = stats[stats$pmax >= 0.95, ], size = 4, 
                  aes(y = pmax, x = '', label = gene_name), 
                  position = position_jitter(width = 0.5, height = 0))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  xlab(label = "")+
  ylab(label = expression(pi))+
  ylim(c(0.5, 1))
```



### Promising hits
Lets visualize the observed data of the genes with promising probability of 
differential gene usage ($\pi > 0.95$). Here we show the gene usage in %.

```{r, fig.height = 5, fig.wide = TRUE}
promising.genes <- stats$gene_name[stats$pmax >= 0.95]

group.ppc <- M$group.ppc$group.ppc
group.ppc <- group.ppc[group.ppc$gene_name %in% promising.genes, ]

point.data <- M$group.ppc$individual.pct.data
point.data <- point.data[point.data$gene_name %in% promising.genes, ]

ggplot()+
  geom_errorbar(data = group.ppc,
                aes(x = gene_name, ymin = ppc.L, 
                    ymax = ppc.H, col = condition),
                position = position_dodge(width = .8), width = 0.75)+
  geom_point(data = point.data,
             aes(x = gene_name, y = gene_usage_pct, col = condition),
             shape = 21, size = 1.5, fill = "black",
             position = position_jitterdodge(jitter.width = 0.15, 
                                             jitter.height = 0, 
                                             dodge.width = 0.8))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  ylab(label = "Gene usage [%]")+
  xlab(label = '')
```


### Promising hits [count]
Lets also visualize the gene usage in raw counts. Bubble size represents total
usage in each repertoire.

```{r, fig.height = 5, fig.wide = TRUE}
promising.genes <- stats$gene_name[stats$pmax >= 0.95]

ggplot(data = viz[viz$gene_name %in% promising.genes, ])+
  geom_point(aes(x = gene_name, y = gene_usage_count, 
                 fill = condition, size = total/10^3, alpha = 0.5), shape = 21,
             position = position_jitterdodge(dodge.width = 0.75,
                                             jitter.width = 0.2))+
  theme_bw(base_size = 12)+
  ylab(label = "Usage [count]")+
  xlab(label = '')+
  theme(legend.position = "top")+
  scale_fill_manual(name = "Condition", values = c("orange", "#6666ff"))+
  scale_size_continuous(name = "N (10^3)", range = c(2, 6))+
  ylab(label = "Gene usage [count]")+
  xlab(label = '')
```





## Comparison with the t-test
Despite the fact that both assumptions made by the t-test are violated, i.e.
(1) the assumption of normally distributed data and (2) homoscedasticity, the
t-test if often used in the literature for differential gene usage analysis. 
To use the t-test, researchers compute the proportions of gene usage. With that, 
they discard data about the sample size. Here, we nevertheless compare our 
results against P-values obtained with a t-test (with assumption of unequal 
variance) to illustrate the pitfalls of such analyses. These statistics are 
automatically computed by IgGeneUsage.

The following figure compares the probability ($\pi$) of differential gene
usage obtained from the Bayesian analysis with P-values obtained from the 
t-test. Legend:

  * y-axis: FDR-corrected P-value shown on -log10 scale. Dashed lines represent 
  the significance levels of 0.05 and 0.01. 
  * x-axis: $\pi$
  
We can notice a few deviations between the two analyses. Lets inspect the 
corresponding genes more closely.

```{r, fig.height=6, fig.wide = TRUE}
ggplot()+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), 
             linetype = "dashed", col = "darkgray")+
  geom_point(data = stats,
             aes(x = pmax, y = -log10(t.test.fdr.pvalue)), size = 2)+
  geom_text_repel(data = stats[stats$pmax >= 0.8, ],
                  aes(x = pmax, y = -log10(t.test.fdr.pvalue), label=gene_name), 
                  size = 4, min.segment.length = 0.1)+
  xlim(0.5, 1)+
  ylab(label = "-log10 (P-value) from t-test [FDR corrected]")+
  xlab(label = expression(pi))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  scale_color_discrete(name = '')
```



### Outliers: IGHV1-58 and IGHV3-72 [%]
Based on the t-test results, the genes IGHV1-58 and IGHV3-72 are top candidates 
having differential usage between the two conditions. Their $\pi$ estimates are
however far from 1. 

Why is that so? Based on the gene usage presented in unit %, we get misleading
impression of differential gene usage. If we consider the sample size (see next
figure), we see why basing the differential gene usage analysis based on the
propoertions is innapropriate.

```{r, fig.height = 6, fig.wide = TRUE}
promising.genes <- c("IGHV1-58", "IGHV3-72")

group.ppc <- M$group.ppc$group.ppc
group.ppc <- group.ppc[group.ppc$gene_name %in% promising.genes, ]

point.data <- M$group.ppc$individual.pct.data
point.data <- point.data[point.data$gene_name %in% promising.genes, ]

ggplot()+
  geom_errorbar(data = group.ppc,
                aes(x = gene_name, ymin = ppc.L, 
                    ymax = ppc.H, col = condition),
                position = position_dodge(width = .8), width = 0.75)+
  geom_point(data = point.data,
             aes(x = gene_name, y = gene_usage_pct, col = condition),
             shape = 21, size = 1.5, fill = "black",
             position = position_jitterdodge(jitter.width = 0.2, 
                                             jitter.height = 0, 
                                             dodge.width = 0.75))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  ylab(label = "Gene usage [%]")+
  xlab(label = '')
```


### Outliers: IGHV1-58 and IGHV3-72 [counts]

```{r, fig.height = 6, fig.wide = TRUE}
promising.genes <- c("IGHV1-58", "IGHV3-72")

ggplot(data = viz[viz$gene_name %in% promising.genes, ])+
  geom_point(aes(x = gene_name, y = gene_usage_count, 
                 fill = condition, size = total/10^3, alpha = 0.5), shape = 21,
             position = position_jitterdodge(dodge.width = 0.75,
                                             jitter.width = 0.2))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  scale_fill_manual(name = "Condition", values = c("orange", "#6666ff"))+
  scale_size_continuous(name = "N (10^3)", range = c(2, 6))+
  ylab(label = "Gene usage [count]")+
  xlab(label = '')
```





## Comparison with the Wilcoxon rank-sum test
Another popular method for differential gene usage analysis is the nonparametric
Wilcoxon rank-sum test (U-test). Its assumption of expecting equal shape is also 
not met by our data. Here, we nevertheless compare our results against P-values 
obtained with the U-test, which are automatically computed by IgGeneUsage.

The following figure compares the probability ($\pi$) of differential gene
usage obtained from the Bayesian analysis with P-values obtained from the 
U-test. Legend:

  * y-axis: FDR-corrected P-value shown on -log10 scale. Dashed lines represent 
  the significance levels of 0.05 and 0.01. 
  * x-axis: $\pi$
  
We observe similar results as in the case of the t-test. We can see that the 
U-test is under-powered to detect statistically significant differences in gene 
usage. This is both due to the low sample size and also due to the correction
for multiple comparison.

```{r, fig.height = 6, fig.wide = TRUE}
ggplot()+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), 
             linetype = "dashed", col = "darkgray")+
  geom_point(data = stats,
             aes(x = pmax, y = -log10(u.test.fdr.pvalue)), size = 2)+
  geom_text_repel(data = stats[stats$pmax >= 0.8, ],
                  aes(x = pmax, y = -log10(u.test.fdr.pvalue), label=gene_name), 
                  size = 3, min.segment.length = 0.1)+
  xlim(0.5, 1)+
  ylab(label = "-log10 (P-value) from U-test [FDR corrected]")+
  xlab(label = expression(pi))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  scale_color_discrete(name = '')
```




# Case Study II

A small example database subset from study evaluating vaccine-induced changes in
B-cell populations [^3]. The IGHV family usage is reported in four B-cell
populations (samples IgM, IgD, IgG and IgA) across two timepoints (conditions
= -1 hour and +7 days). In this case study we investigate the overal effect of
the vaccine on the IGHV family usage.

[^3]: Laserson U and Vigneault F, et al. High-resolution antibody dynamics of
vaccine-induced immune responses. Proc Natl Acad Sci USA. 2014 111:4928-33.



## Input data

```{r}
data(Ig)
```


```{r}
kable(x = head(Ig), row.names = FALSE)
```



## Data visualization
Lets look at the input data with ggplot2.

```{r, fig.height = 6, fig.wide = TRUE}
# we can compute the total number of rearrangements per sample
total.usage <- aggregate(gene_usage_count~sample_id+condition,
                         FUN = sum, data = Ig)
total.usage$total <- total.usage$gene_usage_count
total.usage$gene_usage_count <- NULL

# merge it with the original data
viz <- merge(x = Ig, y = total.usage,
             by = c("sample_id", "condition"),
             all.x = TRUE)

# compute %
viz$gene_usage_pct <- viz$gene_usage_count/viz$total*100

# visualize
ggplot(data = viz)+
  geom_point(aes(x = gene_name, y = gene_usage_pct, fill = condition, 
                 shape = condition), stroke = 0, size = 3,
             position = position_jitterdodge(jitter.width = 0.15, 
                                             dodge.width = 0.5))+
  theme_bw(base_size = 12)+
  ylab(label = "Usage [%]")+
  xlab(label = '')+
  theme(legend.position = "top")+
  scale_fill_manual(name = "Condition", values = c("orange", "#6666ff"))+
  scale_shape_manual(name = "Condition", values = c(21, 22))
```

## Analysis with IgGeneUsage

This is the main method of IgGeneUsage for differential gene usage.

```{r}
# M <- diffUsage(usage.data = Ig,
#                mcmc.warmup = 1000,
#                mcmc.steps = 2500,
#                mcmc.chains = 4,
#                mcmc.cores = 4,
#                hdi.level = 0.95,
#                adapt.delta = 0.95,
#                max.treedepth = 13,
#                dev.model = "src/stan_files/beta_binomial_model.stan")
# save(M, file = "R/dev/saved_temp/M_IG.RData")
M <- get(load(file = "R/dev/alakazam_zibb_model.RData"))
```




## Model checking

### Checking MCMC sampling

  * divergences, tree-depth, energy

```{r}
rstan::check_hmc_diagnostics(M$glm)
```

  * Rhat and Neff

```{r, fig.height = 3, fig.wide = TRUE}
grid.arrange(rstan::stan_rhat(object = M$glm),
             rstan::stan_ess(object = M$glm),
             nrow = 1)
```


### Posterior predictive checks: repertoire-specific

```{r, fig.height = 6, fig.wide = TRUE}
ggplot(data = M$ppc)+
  facet_wrap(facets = ~sample_id, ncol = 4)+
  geom_abline(intercept = 0, slope = 1)+
  geom_errorbar(aes(x = raw, y = ppc.raw.mean, ymin = ppc.raw.L, 
                    ymax = ppc.raw.H), col = "darkgray")+
  geom_point(aes(x = raw, y = ppc.raw.mean, fill = condition), 
             shape = 21, size = 2)+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  scale_x_log10()+
  scale_y_log10()+
  xlab(label = "Observed usage [counts]")+
  ylab(label = "Predicted usage [counts]")
```




### Gene-specific PPC

```{r, fig.height = 6, fig.wide = TRUE}
ggplot(data = M$group.ppc$group.ppc)+
  geom_abline(intercept = 0, slope = 1)+
  geom_errorbar(aes(x = raw.M, ymin = ppc.L, ymax = ppc.H), col = "darkgray")+
  geom_point(aes(x = raw.M, y = ppc.M, fill = condition), 
             shape = 21, size = 2)+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  xlab(label = "Observed usage [%]")+
  ylab(label = "Predicted usage [%]")
```



### Leave-one-out comparison
  * with R-package loo

```{r}
loo::loo(x = loo::extract_log_lik(stanfit = M$glm))
```


## Results

```{r, fig.height = 4, fig.wide = TRUE}
# format data
stats <- M$glm.summary
stats <- stats[order(abs(stats$effect_mean), decreasing = F), ]
stats$gene_fac <- factor(x = stats$gene_name,
                         levels = stats$gene_name)

stats <- merge(x= stats, y = M$test.stats, by = "gene_name")

ggplot()+
  geom_point(data = stats[stats$pmax < 0.95, ], aes(y = pmax, x = ''),
             position = position_jitter(width = 0.25, height = 0))+
  # geom_text(data = stats[stats$pmax >= 0.95, ], size = 3, 
  #                 aes(y = pmax, x = '', label = gene_name), 
  #                 position = position_jitter(width = 0.5, height = 0))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  xlab(label = "")+
  ylab(label = expression(pi))+
  ylim(c(0.5, 1))
```





## Comparison with the t-test
We again compare compare the probability ($\pi$) of differential gene
usage obtained from the Bayesian analysis, with P-values obtained from the 
t-test. Legend:

  * y-axis: FDR-corrected P-value shown on -log10 scale. Dashed lines represent 
  the significance levels of 0.05 and 0.01. 
  * x-axis: $\pi$
  
We can notice a few deviations between the two analyses. Lets inspect the 
corresponding genes more closely.

```{r, fig.height = 5, fig.wide = TRUE}
ggplot(data = stats)+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), 
             linetype = "dashed", col = "darkgray")+
  geom_point(aes(x = pmax, y = -log10(t.test.fdr.pvalue)), size = 2)+
  geom_text_repel(aes(x = pmax, y = -log10(t.test.fdr.pvalue), 
                label = gene_name), size = 3)+
  xlim(0.5, 1)+
  ylab(label = "-log10 (P-value) from t-test [FDR corrected]")+
  xlab(label = expression(pi))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  scale_color_discrete(name = '')
```



### Outliers: IGHV5

```{r, fig.height = 5, fig.wide = TRUE}
promising.genes <- unique(M$usage.data$gene_names)

group.ppc <- M$group.ppc$group.ppc
group.ppc <- group.ppc[group.ppc$gene_name %in% promising.genes, ]

point.data <- M$group.ppc$individual.pct.data
point.data <- point.data[point.data$gene_name %in% promising.genes, ]

ggplot()+
  geom_errorbar(data = group.ppc,
                aes(x = gene_name, ymin = ppc.L, 
                    ymax = ppc.H, col = condition),
                position = position_dodge(width = .8), width = 0.75)+
  geom_point(data = point.data,
             aes(x = gene_name, y = gene_usage_pct, col = condition),
             shape = 21, size = 1.5, fill = "black",
             position = position_jitterdodge(jitter.width = 0.15, 
                                             jitter.height = 0, 
                                             dodge.width = 0.8))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")
```


### Outliers: IGHV5 (original data)

```{r, fig.height = 5, fig.wide = TRUE}
ggplot(data = viz[viz$gene_name %in% promising.genes, ])+
  geom_point(aes(x = gene_name, y = gene_usage_count, 
                 fill = condition, size = total/10^2, alpha = 0.5), shape = 21,
             position = position_jitterdodge(dodge.width = 0.5,
                                             jitter.width = 0.15))+
  theme_bw(base_size = 12)+
  ylab(label = "Usage [count]")+
  xlab(label = '')+
  theme(legend.position = "top")+
  scale_fill_manual(name = "Condition", values = c("orange", "#6666ff"))+
  scale_size_continuous(name = "N (10^2)", range = c(2, 6))
```





## Comparison with the Wilcoxon rank-sum test (U-test)
Both the U-test and IgGeneUsage report no differential usage.

```{r, fig.height = 5, fig.wide = TRUE}
ggplot(data = stats)+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), 
             linetype = "dashed", col = "darkgray")+
  geom_point(aes(x = pmax, y = -log10(u.test.fdr.pvalue)), size = 2)+
  geom_text_repel(aes(x = pmax, y = -log10(u.test.fdr.pvalue), 
                label = gene_name), size = 4)+
  xlim(0.5, 1)+
  ylab(label = "-log10 (P-value) from U-test [FDR corrected]")+
  xlab(label = expression(pi))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  scale_color_discrete(name = '')
```







# Case Study III

The main purpose of using IgGeneUsage is to detect differential gene usage. 
However, the tool can in principle be be used to tackle all tasks that fall 
under the scope of binomial regression. 

In the third case study we explain how to detect differential usage of net CDR3 
sequence charge between repertoires belonging to two biological conditions. For 
this purpose we use CDR3 sequence data of human T-cells (TRB-chain) downloaded 
from VDJdb [^4]. Each CDR3 sequence is annotated to a specific viral epitope. 
Here we focus on the two viruses InfluenzaA and CSM, which represent our 
biological conditions. We consider the different data sources (publications) 
as different repertoires.

To compute the net sequence charge, we consider the amino acids K, R and H as 
+1 charged, while D and E as -1 charged. Thus, we computed the net charge of a 
CDR3 sequence by adding up the individual residue charges. 

[^4]: https://vdjdb.cdr3.net/


## Input data

```{r}
data(CDR3_Epitopes)
```



```{r}
kable(x = head(CDR3_Epitopes), row.names = FALSE)
```




## Data visualization

Lets look at the input data with ggplot2.

```{r, fig.height = 5, fig.wide = TRUE}
# we can compute the total number of rearrangements per sample
total.usage <- aggregate(gene_usage_count~sample_id+condition,
                         FUN = sum, data = CDR3_Epitopes)
total.usage$total <- total.usage$gene_usage_count
total.usage$gene_usage_count <- NULL

# merge it with the original data
viz <- merge(x = CDR3_Epitopes, y = total.usage,
             by = c("sample_id", "condition"),
             all.x = TRUE)

# compute %
viz$gene_usage_pct <- viz$gene_usage_count/viz$total*100

# visualize
ggplot(data = viz)+
  geom_point(aes(x = as.numeric(gene_name), y = gene_usage_pct, fill = condition, 
                 shape = condition), stroke = 0, size = 2, alpha = 0.5,
             position = position_jitterdodge(jitter.width = 0.15, 
                                             dodge.width = 0.5))+
  theme_bw(base_size = 12)+
  ylab(label = "Usage [%]")+
  xlab(label = 'Net Charge')+
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4))+
  scale_fill_manual(name = "Condition", values = c("orange", "#6666ff"))+
  scale_shape_manual(name = "Condition", values = c(21, 22))+
  scale_x_continuous(breaks = -7:7, labels = -7:7)
```



## Analysis with IgGeneUsage

```{r}
# M <- diffUsage(usage.data = CDR3_Epitopes,
#                mcmc.warmup = 1000,
#                mcmc.steps = 2500,
#                mcmc.chains = 4,
#                mcmc.cores = 4,
#                hdi.level = 0.95,
#                adapt.delta = 0.95,
#                max.treedepth = 13,
#                dev.model = "src/stan_files/zibb_model.stan")

# save(M, file = "R/dev/epitopes_zibb_model.RData")
M <- get(load(file = "R/dev/epitopes_zibb_model.RData"))
```




## Model checking

### MCMC sampling

  * divergences, tree-depth, energy

```{r}
rstan::check_hmc_diagnostics(M$glm)
```

  * Rhat and Neff

```{r, fig.height = 3, fig.wide = TRUE}
grid.arrange(rstan::stan_rhat(object = M$glm),
             rstan::stan_ess(object = M$glm),
             nrow = 1)
```


### Posterior predictive checks: repertoire-specific

```{r, fig.height = 8, fig.wide = TRUE}
ggplot(data = M$ppc)+
  facet_wrap(facets = ~sample_id, ncol = 5)+
  geom_abline(intercept = 0, slope = 1)+
  geom_errorbar(aes(x = raw, y = ppc.raw.mean, ymin = ppc.raw.L, 
                    ymax = ppc.raw.H), col = "darkgray")+
  geom_point(aes(x = raw, y = ppc.raw.mean, fill = condition), 
             shape = 21, size = 2.5)+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  scale_x_log10()+
  scale_y_log10()+
  xlab(label = "Observed usage [counts]")+
  ylab(label = "Predicted usage [counts]")
```




### Posterior predictive checks: overall differential usage

```{r, fig.height = 4, fig.wide = TRUE}
ggplot(data = M$group.ppc$group.ppc)+
  geom_abline(intercept = 0, slope = 1)+
  geom_errorbar(aes(x = raw.M, ymin = ppc.L, ymax = ppc.H), col = "darkgray")+
  geom_point(aes(x = raw.M, y = ppc.M, fill = condition), 
             shape = 21, size = 2)+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  xlab(label = "Observed usage [%]")+
  ylab(label = "Predicted usage [%]")
```



### Leave-one-out
  * with R-package loo

```{r}
loo::loo(x = loo::extract_log_lik(stanfit = M$glm))
```


## Results

```{r, fig.height = 4, fig.wide = TRUE}
stats <- M$glm.summary
stats <- stats[order(abs(stats$effect_mean), decreasing = F), ]
stats$gene_fac <- factor(x = stats$gene_name, levels = stats$gene_name)
stats <- merge(x= stats, y = M$test.stats, by = "gene_name")

ggplot()+
  geom_point(data = stats[stats$pmax < 0.95, ], aes(y = pmax, x = ''),
             position = position_jitter(width = 0.25, height = 0))+
  geom_text(data = stats[stats$pmax >= 0.95, ], size = 5, 
                  aes(y = pmax, x = '', label = gene_name), 
                  position = position_jitter(width = 0.5, height = 0))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  xlab(label = "")+
  ylab(label = expression(pi))+
  ylim(c(0.5, 1))
```





## Comparison with the t-test
We again compare $\pi$ estimates from IgGeneUsage with the P-values obtained 
from the t-test. Legend:

  * y-axis: FDR-corrected p-value shown on -log10 scale. Dashed lines represent 
  the significance levels of 0.05 and 0.01. 
  * x-axis: $\pi$

```{r, fig.height=5, fig.wide = TRUE}
ggplot(data = stats)+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), 
             linetype = "dashed", col = "darkgray")+
  geom_point(aes(x = pmax, y = -log10(t.test.fdr.pvalue)), size = 1.5)+
  geom_text_repel(aes(x = pmax, y = -log10(t.test.fdr.pvalue), 
                label = gene_name), size = 5)+
  xlim(0.5, 1)+
  ylab(label = "-log10 (P-value) from t-test [FDR corrected]")+
  xlab(label = expression(pi))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  scale_color_discrete(name = '')
```








## Comparison with the Wilcoxon rank-sum test (U-test)
The U-test finds differential usage in net chage group = 1. IgGeneUsage is
more conservative, which makes sense when we take the sample sizes into
consideration (see below).

```{r, fig.height=5, fig.wide = TRUE}
ggplot(data = stats)+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), 
             linetype = "dashed", col = "darkgray")+
  geom_point(aes(x = pmax, y = -log10(u.test.fdr.pvalue)), size = 1.5)+
  geom_text_repel(aes(x = pmax, y = -log10(u.test.fdr.pvalue), 
                label = gene_name), size = 5)+
  xlim(0.5, 1)+
  ylab(label = "-log10(p-value) from U-test [FDR corrected]")+
  xlab(label = expression(pi))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  scale_color_discrete(name = '')
```




### Net charge usage [in %] groups and posterior means (and 95% HDI of mean) 

```{r, fig.height=5, fig.wide = TRUE}
promising.genes <- M$usage.data$gene_names

group.ppc <- M$group.ppc$group.ppc
group.ppc <- group.ppc[group.ppc$gene_name %in% promising.genes, ]

point.data <- M$group.ppc$individual.pct.data
point.data <- point.data[point.data$gene_name %in% promising.genes, ]

ggplot()+
  geom_errorbar(data = group.ppc,
                aes(x = gene_name, ymin = ppc.L, 
                    ymax = ppc.H, col = condition),
                position = position_dodge(width = .8), width = 0.75)+
  geom_point(data = point.data,
             aes(x = gene_name, y = gene_usage_pct, col = condition),
             shape = 21, size = 1.5, fill = "black",
             position = position_jitterdodge(jitter.width = 0.2, 
                                             jitter.height = 0, 
                                             dodge.width = 0.75))+
  theme_bw(base_size = 12)+
  theme(legend.position = "top")+
  ylab(label = "Gene usage [%]")+
  xlab(label = '')
```


### Net charge usage [counts] groups and posterior means (and 95% HDI of mean) 
Lets also visualize the gene usage in raw counts. Bubble size represents total
usage in each repertoire.

```{r, fig.height = 5, fig.wide = TRUE}
promising.genes <- M$usage.data$gene_names

ggplot(data = viz[viz$gene_name %in% promising.genes, ])+
  geom_point(aes(x = gene_name, y = gene_usage_count, 
                 fill = condition, size = total/10^3, alpha = 0.5), shape = 21,
             position = position_jitterdodge(dodge.width = 0.75,
                                             jitter.width = 0.2))+
  theme_bw(base_size = 12)+
  ylab(label = "Usage [count]")+
  xlab(label = '')+
  theme(legend.position = "top")+
  scale_fill_manual(name = "Condition", values = c("orange", "#6666ff"))+
  scale_size_continuous(name = "N (10^3)", range = c(2, 6))+
  ylab(label = "Gene usage [count]")+
  xlab(label = '')+
  scale_y_log10()
```
