---
title: "How to use IgGeneUsage"
author: "SK"
date: "June 25, 2019"
output: 
  html_document: 
    highlight: tango
    theme: journal
    toc: yes
---


```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = FALSE, 
                      warning = FALSE, message = FALSE)
```



```{r}
require(ggplot2)
require(gridExtra)
require(ggrepel)
require(rstan)
rstan_options(auto_write = TRUE)

source(file = "R/Util.R")
source(file = "R/Usage.R")
```





# Introduction
The analysis of immune repertoires (B-cell receptors and T-cell receptors) 
involves the quantification of different repertoire features (e.g. CDR3 length,
clonal expansion, diversity, gene usage, etc.). Sometimes our intention is to 
discover distruptions in repertoire features between biological conditions (e.g.
healthy vs infected, normal vs tumor). IgGeneUsage can help you quantify such
disruptions in gene usage between two conditions.

## Usage
All you need to carry out your analysis with IgGeneUsage is a data frame with 
the following 4 columns:

  1. sample_id: character identifier of the sample/repertoire (e.g. Patient-1)
  2. condition: character identifier of the condition to which each repertoire 
  belongs (e.g. healthy or tumor)
  3. gene_name: specific gene name (e.g. IGHV1-10 or family TRV1)
  4. gene_usage_count: numeric (count) of rearrangements related to columns 1-3

Important remark: The sum of the gene_usage_count (column 4) for a given sample 
(sample_id) should be equal to the total number of rearrangements in that 
repertoire. 
The remaining parameters provided to IgGeneUsage help the MCMC sampler. The
default settings should suffice for most analyses.

# Case Studies
## Case Study I
The IgGeneUsage package comes with built-in datasets of gene usage. In the first 
case study we will evaluate disruptions in the IGHV-segment usage of class 
switched (CS) memory B-cells between HCV+ (N = 22) and healthy individuals 
(N = 7)[^1].

[^1]: Tucci, Felicia A., et al. "Biased IGH VDJ gene repertoire and clonal 
expansions in B cells of chronically hepatitis C virus–infected individuals." 
Blood 131.5 (2018): 546-557.]


### Input data

```{r}

IGHV_HCV <- read.csv(file = "inst/IGHV_HCV_CSM.csv", sep = " ", as.is = T)
IGHV_HCV <- IGHV_HCV[, c("sample_id", "condition", 
                         "gene_name", "gene_usage_count")]
head(IGHV_HCV)
```


### Data visualization
Lets look at the input data with ggplot2.

```{r, fig.width=7, fig.height=4}
# we can compute the total number of rearrangements per sample
total.usage <- aggregate(gene_usage_count~sample_id+condition, 
                         FUN = sum, data = IGHV_HCV)
total.usage$total <- total.usage$gene_usage_count
total.usage$gene_usage_count <- NULL

# merge it with the original data
viz <- merge(x = IGHV_HCV, y = total.usage, 
                  by = c("sample_id", "condition"), 
                  all.x = TRUE)

# compute %
viz$gene_usage_pct <- viz$gene_usage_count/viz$total*100

# for this visualization select genes with >1% usage in all samples
viz$bit <- viz$gene_usage_pct < 1
bit.m <- table(viz$gene_name, viz$bit)
viz <- viz[!viz$gene_name %in% names(which(bit.m[, "TRUE"] == 29)), ]

# visualize
ggplot(data = viz)+
  geom_point(aes(x = gene_name, y = gene_usage_pct, 
                 fill = condition, shape = condition),
             position = position_dodge(width = 1), stroke = 0)+
  theme_bw(base_size = 9)+
  ylab(label = "Usage [%]")+
  xlab(label = '')+
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4))+
  scale_fill_manual(name = "Condition", values = c("orange", "#6666ff"))+
  scale_shape_manual(name = "Condition", values = c(21, 22))
```


### Analysis with IgGeneUsage
This is the main method of IgGeneUsage for differential gene usage.

```{r}
#   M <- diffUsage(usage.data = IGHV_HCV,
#                              mcmc.warmup = 1000,
#                              mcmc.steps = 2500,
#                              mcmc.chains = 4,
#                              mcmc.cores = 4,
#                              hdi.level = 0.95,
#                              adapt.delta = 0.95,
#                              max.treedepth = 13,
#                              dev.model = stan.files[m])
M <- get(load(file = "R/dev/ighv_hcv_beta_binomial_model.RData"))
```




### Model checking
Do not blindly trust the model $\rightarrow$ check it extensively. 

  * Checklist of successful MCMC sampling:
      * no divergences
      * no additional warnings from rstan
      * Rhat < 1.05 (see object glm.stats)
      * sufficiently high Neff
  * Checklist for valid model:
      * posterior predictive check reveal decent predictive accuracy on 
        already observed data (and also new data)
      * leave-one-out analysis with R-package loo (most k values should be low)

#### Checking MCMC sampling

  * divergences, tree-depth, energy
  
```{r}
rstan::check_hmc_diagnostics(M$glm)
```

  * Rhat and Neff

```{r, fig.width=7, fig.height=3}
grid.arrange(rstan::stan_rhat(object = M$glm),
             rstan::stan_ess(object = M$glm),
             nrow = 1)
```


#### Repertoire-specific PPC

```{r, fig.height=6, fig.width=8}
ggplot(data = M$ppc)+
  facet_wrap(facets = ~sample_id, ncol = 7)+
  geom_abline(intercept = 0, slope = 1)+
  geom_point(aes(x = raw, y = ppc.raw.mean, fill = condition), shape = 21)+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```




#### Gene-specific PPC

```{r, fig.width=3, fig.height=3}
ggplot(data = M$group.ppc$group.ppc)+
  geom_abline(intercept = 0, slope = 1)+
  geom_errorbar(aes(x = raw.M, ymin = ppc.L, ymax = ppc.H))+
  geom_point(aes(x = raw.M, y = ppc.M, fill = condition), shape = 21)+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```



#### Leave-one-out comparison
  * with R-package loo
  
```{r}
loo::loo(x = loo::extract_log_lik(stanfit = M$glm))
```


### Results
This is the main result of IgGeneUsage $\rightarrow$ disruption effects on each
gene due to the biological condition. Legend:

  * Y-axis: all genes
  * X-axis: effect (positive = increase in usage in HCV, negative = decrease)
  * dashed-line: null effect
  * points: mean estimated effect
  * confidence intervals: 95% highest density interval (HDI) of effect

For simple interpretation of the results, we color differently the effects 
based on whether they include/exclude the null-effect within their 95% HDI.

```{r, fig.height=8, fig.width=5}
# format data
stats <- M$glm.summary
stats$effect_col <- ifelse(test = stats$effect_L <= 0 & stats$effect_H >= 0,
                           yes = "include", no = "exclude")
stats <- stats[order(abs(stats$effect_mean), decreasing = F), ]
stats$gene_fac <- factor(x = stats$gene_name, 
                         levels = stats$gene_name)


# visualize disruption effects on each gene
ggplot(data = stats)+
  geom_vline(xintercept = 0, linetype = "dashed", col = "darkgray")+
  geom_errorbarh(aes(y = gene_fac, xmin = effect_L, 
                     xmax = effect_H, col = effect_col))+
  geom_point(aes(y = gene_fac, x = effect_mean))+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```



#### Some interesting genes
There are a few genes in which IgGeneUsage found some disruption between the 
two conditions. Here we visualize the data and the mean probability (error 
bars showing 95% HDI of mean) for each gene and condition. Notice that the 
data violates the assumptions made by the t-test and the non-parameteric
Wilcoxon rank-sum test (U-test). Both methods are frequently used in the 
literature for the analysis of differential gene usage.

```{r, fig.width=6, fig.height=3.5}
group.ppc <- M$group.ppc$group.ppc
group.ppc <- group.ppc[group.ppc$gene_name %in% 
                         stats$gene_name[stats$effect_col == "exclude"], ]

point.data <- M$group.ppc$individual.pct.data
point.data <- point.data[point.data$gene_name %in% 
                           stats$gene_name[stats$effect_col == "exclude"], ]

ggplot()+
  geom_errorbar(data = group.ppc,
                aes(x = gene_name, ymin = ppc.L, 
                    ymax = ppc.H, col = condition),
                position = position_dodge(width = .8), width = 0.75)+
  geom_point(data = point.data,
             aes(x = gene_name, y = gene_usage_pct, col = condition),
             shape = 21, size = 1.5, fill = "black",
             position = position_jitterdodge(jitter.width = 0.15, 
                                             jitter.height = 0, 
                                             dodge.width = 0.8))+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```



#### Comparison with the t-test
We nevertheless compare our results against P-values obtained with a t-test 
(with assumption of unequal variance), which are automatically estimated by
IgGeneUsage based on the data of proportions (not the raw usages). 

Important remark: Please have in mind that this step, although necessary if we 
want to apply the t-test, discards very useful data about the sample size. The 
dashed lines represent significance levels of 0.05 and 0.01.

```{r, fig.width=6, fig.height=5}
stats <- merge(x= stats, y = M$test.stats, by = "gene_name")

ggplot(data = stats)+
  geom_text_repel(aes(x = pmax, y = -log10(t.test.fdr.pvalue), 
                label = gene_name, col = effect_col), size = 2.5)+
  xlim(0.5, 1)+
  ylab(label = "-log10(p-value) from t-test [FDR corrected]")+
  xlab(label = "pmax")+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), linetype = "dashed")+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```


#### Outliers
Lets look at the genes in which we see discrepancy between the two analyses.

```{r, fig.width=6, fig.height=3}
group.ppc <- M$group.ppc$group.ppc
group.ppc <- group.ppc[group.ppc$gene_name %in% c("IGHV1-58", 
                                                  "IGHV3-72"), ]

point.data <- M$group.ppc$individual.pct.data
point.data <- point.data[point.data$gene_name %in% c("IGHV1-58", 
                                                     "IGHV3-72"), ]

ggplot()+
  geom_errorbar(data = group.ppc,
                aes(x = gene_name, ymin = ppc.L, ymax = ppc.H, col = condition),
                position = position_dodge(width = .8), width = 0.75)+
  geom_point(data = point.data,
             aes(x = gene_name, y = gene_usage_pct, col = condition),
             shape = 21, size = 1.5, fill = "black",
             position = position_jitterdodge(jitter.width = 0.15, 
                                             jitter.height = 0, 
                                             dodge.width = 0.8))+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")

rm(point.data, group.ppc)
```



#### Comparison with the Wilcoxon rank-sum test (u-test)
Let us also compare our results against P-values obtained with the u-test. The
outcome is similar as in the case of the t-test. These results show that the 
u-test is under-powered to detect statistically significant differences in gene 
usage. This is both due to the low sample size and also due to the correction
for multiple comparison.

Important remark: The u-test makes assumption about the data as well (equal 
shape). Are they satisfied?

```{r, fig.width=5, fig.height=5}
ggplot(data = stats)+
  geom_text_repel(aes(x = pmax, y = -log10(u.test.fdr.pvalue), 
                label = gene_name, col = effect_col), size = 2.5)+
  xlim(0.5, 1)+
  ylab(label = "-log10(p-value) from u-test [FDR corrected]")+
  xlab(label = "pmax")+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), linetype = "dashed")+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```



## Case Study II
The second case study considers unpublished data to evaluate disruptions in the 
IGHV-segment usage of memory B-cells between two samples 1) HIV+ bult (N = 7) 
and HIV+ gp120 (N = 6).

### Input data

```{r}
# data(IGHV_HIV)
IGHV_HIV <- read.csv(file = "inst/IGHV_HIV.csv", sep = " ", as.is = T)
IGHV_HIV <- IGHV_HIV[, c("sample_id", "condition", 
                         "gene_name", "gene_usage_count")]
head(IGHV_HIV)
```


### Data visualization
Lets look at the input data with ggplot2.

```{r, fig.width=7, fig.height=4}
# we can compute the total number of rearrangements per sample
total.usage <- aggregate(gene_usage_count~sample_id+condition, 
                         FUN = sum, data = IGHV_HIV)
total.usage$total <- total.usage$gene_usage_count
total.usage$gene_usage_count <- NULL

# merge it with the original data
viz <- merge(x = IGHV_HIV, y = total.usage, 
                  by = c("sample_id", "condition"), 
                  all.x = TRUE)

# compute %
viz$gene_usage_pct <- viz$gene_usage_count/viz$total*100

# for this visualization select genes with >1% usage in all samples
viz$bit <- viz$gene_usage_pct < 1
bit.m <- table(viz$gene_name, viz$bit)
viz <- viz[!viz$gene_name %in% names(which(bit.m[, "TRUE"] == 13)), ]

# visualize
ggplot(data = viz)+
  geom_point(aes(x = gene_name, y = gene_usage_pct, 
                 fill = condition, shape = condition),
             position = position_dodge(width = 1), stroke = 0)+
  theme_bw(base_size = 9)+
  ylab(label = "Usage [%]")+
  xlab(label = '')+
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4))+
  scale_fill_manual(name = "Condition", values = c("orange", "#6666ff"))+
  scale_shape_manual(name = "Condition", values = c(21, 22))
```

### Analysis with IgGeneUsage
This is the main method of IgGeneUsage for differential gene usage.

```{r}
# M <- diffUsage(usage.data = IGHV_HIV,
#                mcmc.warmup = 1000,
#                mcmc.steps = 2500,
#                mcmc.chains = 4,
#                mcmc.cores = 4,
#                hdi.level = 0.95,
#                adapt.delta = 0.95,
#                max.treedepth = 13,
#                dev.model = stan.files[m])
M <- get(load(file = "R/dev/ighv_hiv_beta_binomial_model.RData"))
```




### Model checking
Do not blindly trust the model $\rightarrow$ check it extensively. 

  * Checklist of successful MCMC sampling:
      * no divergences
      * no additional warnings from rstan
      * Rhat < 1.05 (see object glm.stats)
      * sufficiently high Neff
  * Checklist for valid model:
      * posterior predictive check reveal decent predictive accuracy on 
        already observed data (and also new data)
      * leave-one-out analysis with R-package loo (most k values should be low)

#### Checking MCMC sampling

  * divergences, tree-depth, energy
  
```{r}
rstan::check_hmc_diagnostics(M$glm)
```

  * Rhat and Neff

```{r, fig.width=7, fig.height=3}
grid.arrange(rstan::stan_rhat(object = M$glm),
             rstan::stan_ess(object = M$glm),
             nrow = 1)
```


#### Repertoire-specific PPC

```{r, fig.height=5, fig.width=7}
ggplot(data = M$ppc)+
  facet_wrap(facets = ~sample_id, ncol = 5)+
  geom_abline(intercept = 0, slope = 1)+
  geom_point(aes(x = raw, y = ppc.raw.mean, fill = condition), shape = 21)+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```




#### Gene-specific PPC

```{r, fig.width=3, fig.height=3}
ggplot(data = M$group.ppc$group.ppc)+
  geom_abline(intercept = 0, slope = 1)+
  geom_errorbar(aes(x = raw.M, ymin = ppc.L, ymax = ppc.H))+
  geom_point(aes(x = raw.M, y = ppc.M, fill = condition), shape = 21)+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```



#### Leave-one-out comparison
  * with R-package loo
  
```{r}
loo::loo(x = loo::extract_log_lik(stanfit = M$glm))
```


### Results
This is the main result of IgGeneUsage $\rightarrow$ disruption effects on each
gene due to the biological condition. Legend:

  * Y-axis: all genes
  * X-axis: effect (positive = increase in usage in HCV, negative = decrease)
  * dashed-line: null effect
  * points: mean estimated effect
  * confidence intervals: 95% highest density interval (HDI) of effect

For simple interpretation of the results, we color differently the effects based
on whether they include/exclude the null-effect within their 95% HDI.

```{r, fig.height=8, fig.width=5}
# format data
stats <- M$glm.summary
stats$effect_col <- ifelse(test = stats$effect_L <= 0 & stats$effect_H >= 0,
                           yes = "include", no = "exclude")
stats <- stats[order(abs(stats$effect_mean), decreasing = F), ]
stats$gene_fac <- factor(x = stats$gene_name, 
                         levels = stats$gene_name)
stats <- merge(x= stats, y = M$test.stats, by = "gene_name")

# visualize disruption effects on each gene
ggplot(data = stats)+
  geom_vline(xintercept = 0, linetype = "dashed", col = "darkgray")+
  geom_errorbarh(aes(y = gene_fac, xmin = effect_L, 
                     xmax = effect_H, col = effect_col))+
  geom_point(aes(y = gene_fac, x = effect_mean))+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```



#### Some interesting genes
There are a few genes in which IgGeneUsage found some disruption between the two
conditions. Here we visualize the data and the mean probability (error bars 
showing 95% HDI of mean) for each gene and condition. Notice that the data 
seriously violates most assumptions made by the t-test and the nonparameteric
Wilcoxon rank-sum test (U-test) which are most frequently used in the literature
for comparison of differential gene usage.

```{r, fig.width=6, fig.height=3.5}
group.ppc <- M$group.ppc$group.ppc
group.ppc <- group.ppc[group.ppc$gene_name %in% c("IGHV4-4",
                                                  "IGHV3-30-3",
                                                  "IGHV4-61",
                                                  "IGHV5-51"), ]

point.data <- M$group.ppc$individual.pct.data
point.data <- point.data[point.data$gene_name %in% c("IGHV4-4",
                                                     "IGHV3-30-3",
                                                     "IGHV4-61",
                                                     "IGHV5-51"), ]
ggplot()+
  geom_errorbar(data = group.ppc,
                aes(x = gene_name, ymin = ppc.L, 
                    ymax = ppc.H, col = condition),
                position = position_dodge(width = .8), width = 0.75)+
  geom_point(data = point.data,
             aes(x = gene_name, y = gene_usage_pct, col = condition),
             shape = 21, size = 1.5, fill = "black",
             position = position_jitterdodge(jitter.width = 0.15, 
                                             jitter.height = 0, 
                                             dodge.width = 0.8))+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```



#### Comparison with the t-test
We nevertheless compare our results against P-values obtained with a t-test 
(with assumption of unequal variance), which are automatically estimated by
IgGeneUsage based on the data of proportions (not the raw usages). 

Important remark: Please have in mind that this step, although necessary if we 
want to apply the t-test, discards very useful data about the sample size. The 
dashed lines represent significance levels of 0.05 and 0.01.

```{r, fig.width=6, fig.height=5}
ggplot(data = stats)+
  geom_text_repel(aes(x = pmax, y = -log10(t.test.fdr.pvalue), 
                label = gene_name, col = effect_col), size = 2.5)+
  xlim(0.5, 1)+
  ylab(label = "-log10(p-value) from t-test [FDR corrected]")+
  xlab(label = "pmax")+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), linetype = "dashed")+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```





#### Comparison with the Wilcoxon rank-sum test (u-test)
Let us also compare our results against P-values obtained with the u-test. The
outcome is similar as in the case of the t-test. These results show that the 
u-test is under-powered to detect statistically significant differences in gene 
usage. This is both due to the low sample size and also due to the correction
for multiple comparison.

Important remark: The u-test makes assumption about the data as well (equal 
shape). Are they satisfied?

```{r, fig.width=5, fig.height=5}
ggplot(data = stats)+
  geom_text_repel(aes(x = pmax, y = -log10(u.test.fdr.pvalue), 
                label = gene_name, col = effect_col), size = 2.5)+
  xlim(0.5, 1)+
  ylab(label = "-log10(p-value) from u-test [FDR corrected]")+
  xlab(label = "pmax")+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), linetype = "dashed")+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```





## Case Study III
The third case study evaluates the relationships between the T-cell receptors 
(TCR) of T cells within tumors (N = 6), in the surrounding tissues (N = 6) and 
in draining lymph nodes (N = 6) [^2].

[^2]: Wang, Ting, et al. "The different T-cell receptor repertoires in breast 
cancer tumors, draining lymph nodes, and adjacent tissues." Cancer immunology 
research 5.2 (2017): 148-156.

### Input data

```{r}
# data(TRV_Cancer)
TRV_Cancer <- read.csv(file = "inst/TCR_Cancer.csv", sep = ";", as.is = T)
TRV_Cancer <- TRV_Cancer[TRV_Cancer$GroupInformation %in% c("Tumors", 
                                                            "None Tumor"), ]
TRV_Cancer <- TRV_Cancer[which(regexpr(pattern = "TRBV", 
                                       text = TRV_Cancer$ReferenceName) != -1),]

TRV_Cancer$sample_id <- TRV_Cancer$SampleID
TRV_Cancer$condition <- TRV_Cancer$GroupInformation
TRV_Cancer$gene_name <- TRV_Cancer$ReferenceName
TRV_Cancer$gene_usage_count <- TRV_Cancer$UsageNumber

TRV_Cancer <- TRV_Cancer[, c("sample_id", "condition", 
                             "gene_name", "gene_usage_count")]
head(TRV_Cancer)
```


### Data visualization
Lets look at the input data with ggplot2.

```{r, fig.width=7, fig.height=4}
# we can compute the total number of rearrangements per sample
total.usage <- aggregate(gene_usage_count~sample_id+condition, 
                         FUN = sum, data = TRV_Cancer)
total.usage$total <- total.usage$gene_usage_count
total.usage$gene_usage_count <- NULL

# merge it with the original data
viz <- merge(x = TRV_Cancer, y = total.usage, 
                  by = c("sample_id", "condition"), 
                  all.x = TRUE)

# compute %
viz$gene_usage_pct <- viz$gene_usage_count/viz$total*100

# for this visualization select genes with >1% usage in all samples
viz$bit <- viz$gene_usage_pct < 1
bit.m <- table(viz$gene_name, viz$bit)
viz <- viz[!viz$gene_name %in% names(which(bit.m[, "TRUE"] == 13)), ]

# visualize
ggplot(data = viz)+
  geom_point(aes(x = gene_name, y = gene_usage_pct, 
                 fill = condition, shape = condition),
             position = position_dodge(width = 1), stroke = 0)+
  theme_bw(base_size = 9)+
  ylab(label = "Usage [%]")+
  xlab(label = '')+
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4))+
  scale_fill_manual(name = "Condition", values = c("orange", "#6666ff"))+
  scale_shape_manual(name = "Condition", values = c(21, 22))
```

### Analysis with IgGeneUsage
This is the main method of IgGeneUsage for differential gene usage.

```{r}
M <- diffUsage(usage.data = TRV_Cancer,
               mcmc.warmup = 1000,
               mcmc.steps = 2500,
               mcmc.chains = 4,
               mcmc.cores = 4,
               hdi.level = 0.95,
               adapt.delta = 0.95,
               max.treedepth = 13,
               dev.model = "src/stan_files/beta_binomial_model.stan")
# M <- get(load(file = "R/dev/trv_cancer_beta_binomial_model.RData"))
```




### Model checking
Do not blindly trust the model $\rightarrow$ check it extensively. 

  * Checklist of successful MCMC sampling:
      * no divergences
      * no additional warnings from rstan
      * Rhat < 1.05 (see object glm.stats)
      * sufficiently high Neff
  * Checklist for valid model:
      * posterior predictive check reveal decent predictive accuracy on 
        already observed data (and also new data)
      * leave-one-out analysis with R-package loo (most k values should be low)

#### Checking MCMC sampling

  * divergences, tree-depth, energy
  
```{r}
rstan::check_hmc_diagnostics(M$glm)
```

  * Rhat and Neff

```{r, fig.width=7, fig.height=3}
grid.arrange(rstan::stan_rhat(object = M$glm),
             rstan::stan_ess(object = M$glm),
             nrow = 1)
```


#### Repertoire-specific PPC

```{r, fig.height=5, fig.width=7}
ggplot(data = M$ppc)+
  facet_wrap(facets = ~sample_id, ncol = 5)+
  geom_abline(intercept = 0, slope = 1)+
  geom_point(aes(x = raw, y = ppc.raw.mean, fill = condition), shape = 21)+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```




#### Gene-specific PPC

```{r, fig.width=3, fig.height=3}
ggplot(data = M$group.ppc$group.ppc)+
  geom_abline(intercept = 0, slope = 1)+
  geom_errorbar(aes(x = raw.M, ymin = ppc.L, ymax = ppc.H))+
  geom_point(aes(x = raw.M, y = ppc.M, fill = condition), shape = 21)+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```



#### Leave-one-out comparison
  * with R-package loo
  
```{r}
loo::loo(x = loo::extract_log_lik(stanfit = M$glm))
```


### Results
This is the main result of IgGeneUsage $\rightarrow$ disruption effects on each
gene due to the biological condition. Legend:

  * Y-axis: all genes
  * X-axis: effect (positive = increase in usage in HCV, negative = decrease)
  * dashed-line: null effect
  * points: mean estimated effect
  * confidence intervals: 95% highest density interval (HDI) of effect

For simple interpretation of the results, we color differently the effects based
on whether they include/exclude the null-effect within their 95% HDI.

```{r, fig.height=8, fig.width=5}
# format data
stats <- M$glm.summary
stats$effect_col <- ifelse(test = stats$effect_L <= 0 & stats$effect_H >= 0,
                           yes = "include", no = "exclude")
stats <- stats[order(abs(stats$effect_mean), decreasing = F), ]
stats$gene_fac <- factor(x = stats$gene_name, 
                         levels = stats$gene_name)
stats <- merge(x= stats, y = M$test.stats, by = "gene_name")

# visualize disruption effects on each gene
ggplot(data = stats)+
  geom_vline(xintercept = 0, linetype = "dashed", col = "darkgray")+
  geom_errorbarh(aes(y = gene_fac, xmin = effect_L, 
                     xmax = effect_H, col = effect_col))+
  geom_point(aes(y = gene_fac, x = effect_mean))+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```



#### Some interesting genes
There are a few genes in which IgGeneUsage found some disruption between the two
conditions. Here we visualize the data and the mean probability (error bars 
showing 95% HDI of mean) for each gene and condition. Notice that the data 
seriously violates most assumptions made by the t-test and the nonparameteric
Wilcoxon rank-sum test (U-test) which are most frequently used in the literature
for comparison of differential gene usage.

```{r, fig.width=6, fig.height=3.5}
group.ppc <- M$group.ppc$group.ppc
group.ppc <- group.ppc[group.ppc$gene_name %in% c("IGHV4-4",
                                                  "IGHV3-30-3",
                                                  "IGHV4-61",
                                                  "IGHV5-51"), ]

point.data <- M$group.ppc$individual.pct.data
point.data <- point.data[point.data$gene_name %in% c("IGHV4-4",
                                                     "IGHV3-30-3",
                                                     "IGHV4-61",
                                                     "IGHV5-51"), ]
ggplot()+
  geom_errorbar(data = group.ppc,
                aes(x = gene_name, ymin = ppc.L, 
                    ymax = ppc.H, col = condition),
                position = position_dodge(width = .8), width = 0.75)+
  geom_point(data = point.data,
             aes(x = gene_name, y = gene_usage_pct, col = condition),
             shape = 21, size = 1.5, fill = "black",
             position = position_jitterdodge(jitter.width = 0.15, 
                                             jitter.height = 0, 
                                             dodge.width = 0.8))+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```



#### Comparison with the t-test
We nevertheless compare our results against P-values obtained with a t-test 
(with assumption of unequal variance), which are automatically estimated by
IgGeneUsage based on the data of proportions (not the raw usages). 

Important remark: Please have in mind that this step, although necessary if we 
want to apply the t-test, discards very useful data about the sample size. The 
dashed lines represent significance levels of 0.05 and 0.01.

```{r, fig.width=6, fig.height=5}
ggplot(data = stats)+
  geom_text_repel(aes(x = pmax, y = -log10(t.test.fdr.pvalue), 
                label = gene_name, col = effect_col), size = 2.5)+
  xlim(0.5, 1)+
  ylab(label = "-log10(p-value) from t-test [FDR corrected]")+
  xlab(label = "pmax")+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), linetype = "dashed")+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```





#### Comparison with the Wilcoxon rank-sum test (u-test)
Let us also compare our results against P-values obtained with the u-test. The
outcome is similar as in the case of the t-test. These results show that the 
u-test is under-powered to detect statistically significant differences in gene 
usage. This is both due to the low sample size and also due to the correction
for multiple comparison.

Important remark: The u-test makes assumption about the data as well (equal 
shape). Are they satisfied?

```{r, fig.width=5, fig.height=5}
ggplot(data = stats)+
  geom_text_repel(aes(x = pmax, y = -log10(u.test.fdr.pvalue), 
                label = gene_name, col = effect_col), size = 2.5)+
  xlim(0.5, 1)+
  ylab(label = "-log10(p-value) from u-test [FDR corrected]")+
  xlab(label = "pmax")+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), linetype = "dashed")+
  theme_bw(base_size = 9)+
  theme(legend.position = "top")
```
