---
title: "Model Overview"
author: "SK"
date: "June 25, 2019"
output: 
  html_document: 
    highlight: tango
    theme: journal
    toc: yes
---

```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = FALSE, 
                      warning = FALSE, message = FALSE)
```


```{r}
require(ggplot2)
require(ggforce)
require(gridExtra)
require(ggrepel)
source(file = "R/Util.R")
```



# Regression Models

## B: Binomial
* $Y_{ij} \sim \operatorname{Binomial}(N_{i}, \theta_{ij})$
* $\theta_{ij} = \operatorname{logit^{-1}} (\alpha_{j} + \beta_{ij} \cdot X_{i})$
* $\beta_{ij} \sim \operatorname{Normal}(\gamma_{j}, \gamma_{\sigma})$
* $\gamma_{j} \sim \operatorname{Normal}(\hat{\gamma}, \hat{\gamma}_{\sigma})$
* $\alpha_{j} \sim \operatorname{Normal}(\hat{\alpha}, \hat{\alpha}_{\sigma})$
* $\hat{\gamma} \sim \operatorname{Normal}(0, 5)$
* $\hat{\alpha} \sim \operatorname{Normal}(0, 10)$
* $\gamma_{\sigma}, \hat{\gamma}_{\sigma}, \hat{\alpha}_{\sigma} \sim \operatorname{Cauchy^{+}}(0, 1)$

## BB: Beta-binomial
* $Y_{ij} \sim \operatorname{Beta-Binomial}(N_{i}, \theta_{ij} \cdot \phi, (1 - \theta_{ij}) \cdot \phi)$
* $\theta_{ij} = \operatorname{logit^{-1}} (\alpha_{j} + \beta_{ij} \cdot X_{i})$
* $\beta_{ij} \sim \operatorname{Normal}(\gamma_{j}, \gamma_{\sigma})$
* $\gamma_{j} \sim \operatorname{Normal}(\hat{\gamma}, \hat{\gamma}_{\sigma})$
* $\alpha_{j} \sim \operatorname{Normal}(\hat{\alpha}, \hat{\alpha}_{\sigma})$
* $\hat{\gamma} \sim \operatorname{Normal}(0, 5)$
* $\hat{\alpha} \sim \operatorname{Normal}(0, 10)$
* $\gamma_{\sigma}, \hat{\gamma}_{\sigma}, \hat{\alpha}_{\sigma} \sim \operatorname{Cauchy^{+}}(0, 1)$
* $\phi \sim \operatorname{Exponential}(\tau)$
* $\tau \sim \operatorname{Gamma}(3, 0.1)$

## ZIBB: Zero-Inflated Beta-Binomial
* $Y_{ij} \sim $
    * $0$, with prob. $\pi$
    * $\operatorname{Beta-Binomial}(N_{i}, \theta_{ij} \cdot \phi, (1 - \theta_{ij}) \cdot \phi)$, with prob. $1-\pi$
* $\theta_{ij} = \operatorname{logit^{-1}} (\alpha_{j} + \beta_{ij} \cdot X_{i})$
* $\beta_{ij} \sim \operatorname{Normal}(\gamma_{j}, \gamma_{\sigma})$
* $\gamma_{j} \sim \operatorname{Normal}(\hat{\gamma}, \hat{\gamma}_{\sigma})$
* $\alpha_{j} \sim \operatorname{Normal}(\hat{\alpha}, \hat{\alpha}_{\sigma})$
* $\hat{\gamma} \sim \operatorname{Normal}(0, 5)$
* $\hat{\alpha} \sim \operatorname{Normal}(0, 10)$
* $\gamma_{\sigma}, \hat{\gamma}_{\sigma}, \hat{\alpha}_{\sigma} \sim \operatorname{Cauchy^{+}}(0, 1)$
* $\phi \sim \operatorname{Exponential}(\tau)$
* $\tau \sim \operatorname{Gamma}(3, 0.1)$

