---
title: "User Manual: IgGeneUsage"
author: "SK"
date: "Sep 27, 2021"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{User Manual: IgGeneUsage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```



```{r}
require(IgGeneUsage)
require(knitr)
require(ggplot2)
require(ggforce)
require(gridExtra)
require(ggrepel)
require(rstan)
require(reshape2)
rstan_options(auto_write = TRUE)
```



# Introduction
Decoding the properties of immune repertoires is key in understanding the 
response of adaptive immunity to challenges such as viral infection. One 
important property is biases in Ig gene usage between biological conditions 
(e.g. healthy vs turmor). Yet, most analyses for differential gene usage are
performed qualitatively, or with inadequate statistical methods. Here we 
introduce IgGeneUsage, a computational tool for the analysis of differential 
gene usage.


# Input
The input to IgGeneUsage is a data.frame object with usage frequencies for each 
gene of a repertoire that belongs to a particular biological condition. The
usage data.frame has the following 4 columns:

  1. sample_id: identifier of the repertoire (e.g. Patient-1)
  2. condition: identifier of the condition to which each repertoire 
  belongs (e.g. healthy or tumor)
  3. gene_name: specific gene name (e.g. IGHV1-10 or family TRVB1)
  4. gene_usage_count: numeric (count) of usage related to columns 1-3

The sum of all gene usage counts (column 4) for a given repertoire should be 
equal to the total gene usage in that repertoire. 


# Method
IgGeneUsage transforms the provided input in the following way. Given $R$ 
repertoires, each having $G$ genes, IgGeneUsage generates a gene usage matrix 
$Y^{R \times G}$. Row sums in $Y$ define the total usage in each repertoire 
($N$). The design variable $X$ is set to $X = 1$ for repertoires that belong to 
the first condition, and $X = -1$ otherwise.

For the analysis of DGU between two biological conditions, we designed the 
following Bayesian model ($M$) for zero-inflated beta-binomial regression. This
model can fit over-dispersed gene usage data. The immune repertoire data is also
not exhaustive, which leads to misdetection of genes that are systematically 
rearranged at low probability. The zero-inflated component of our model 
accounts for this:

\begin{align}
p(Y_{ij} \mid M) = \begin{cases} 
\kappa + (1 - \kappa) \operatorname{BB}\left(0 \mid N_{i}, \theta_{ij}, \phi 
\right), & \text{if $Y_{ij}$ = 0} \\
(1 - \kappa) \operatorname{BB}\left(Y_{ij} \mid N_{i}, \theta_{ij}, \phi 
\right), & \text{if $Y_{ij}$ > 0}
\end{cases}\\
\theta_{ij} = \operatorname{logit^{-1}}\left(\alpha_{j}+\beta_{ij}X_{i}\right)\\
\beta_{ij}\sim\operatorname{Normal}\left(\gamma_{j},\gamma_{\sigma} \right)\\
\gamma_{j}\sim\operatorname{Normal}\left(\hat{\gamma},\hat{\gamma}_{\sigma}
\right) \\
\alpha_{j}\sim\operatorname{Normal}\left(\hat{\alpha},\hat{\alpha}_{\sigma}
\right) \\
\hat{\gamma} \sim \operatorname{Normal}\left(0, 5\right) \\
\hat{\alpha} \sim \operatorname{Normal}\left(0, 10\right) \\
\gamma_{\sigma}, \hat{\gamma}_{\sigma}, \hat{\alpha}_{\sigma} \sim 
\operatorname{Cauchy^{+}}\left(0, 1\right) \\
\phi \sim \operatorname{Exponential}\left(\tau\right) \\
\tau \sim \operatorname{Gamma}\left(3, 0.1\right) \\
\kappa \sim \operatorname{Beta}\left(1, 3\right)
\end{align}

Model $M$ legend:

  * $i$ and $j$: index of different repertoires and genes, respectively 
  * $\kappa$: zero-inflation probability
  * $\theta$: probability of gene usage
  * $\phi$: dispersion
  * $\alpha$: intercept/baseline gene usage
  * $\beta$: slope/within-repertoire DGU coefficient
  * $\gamma$, $\gamma_{\sigma}$: slope/gene-specific DGU coefficient; 
  standard deviation
  * $\hat{\gamma}$, $\hat{\gamma}_{\sigma}$: mean and standard deviation of 
  the population of gene-specific DGU coefficients
  * $\hat{\alpha}$, $\hat{\alpha}_{\sigma}$: mean and standard deviation of 
  the population of gene-specific baseline usages
  * $\operatorname{BB}$: beta-binomial probability mass function (pmf)
  * $\operatorname{Normal}$: normal probability density function (pdf)
  * $\operatorname{Cauchy^{+}}$: half-Cauchy pdf
  * $\operatorname{Exponential}$: exponential pdf
  * $\operatorname{Gamma}$: gamma pdf
  * $\operatorname{Beta}$: beta pdf
  * $\operatorname{logit^{-1}}$: inverse logistic function

In the output of IgGeneUsage, we report the mean effect size ($\gamma$) and its 
95% highest density interval (HDI). Genes with $\gamma \neq 0$ (e.g. if 95% HDI 
of $\gamma$ excludes 0) are most likely to experience differential usage. 
Additionally, we report the probability of differential gene usage ($\pi$):
\begin{align}
\pi = 2 \cdot \max\left(\int_{\gamma = -\infty}^{0} p(\gamma)\mathrm{d}\gamma, 
\int_{\gamma = 0}^{\infty} p(\gamma)\mathrm{d}\gamma\right) - 1
\end{align}
with $\pi = 1$ for genes with strong differential usage, and $\pi = 0$ for 
genes with negligible differential gene usage. Both metrics are computed based
on the posterior distribution of $\gamma$, and are thus related. We find $\pi$ 
slightly easier to interpret.



# Case Study II
For our second case study we use a dataset from a study evaluating 
vaccine-induced changes in B-cell populations [^5]. The data is publicly 
provided via the R-package alakazam (version 0.2.11). The IGHV family usage is
reported in four B-cell populations (repertoires IgM, IgD, IgG and IgA) across 
two timepoints (conditions = -1 hour vs. +7 days). In this case study we 
investigate the overal effect of the vaccine on the IGHV family usage.

[^5]: Laserson U and Vigneault F, et al. High-resolution antibody dynamics of
vaccine-induced immune responses. Proc Natl Acad Sci USA. 2014 111:4928-33.

## Input data
The data is already formatted such that it can directly be passed as input to
IgGeneUsage. Lets load the data, and to inspect its content.

```{r}
data(Ig, package = "IgGeneUsage")
```

```{r}
kable(x = head(Ig), row.names = FALSE)
```



## Data visualization
Lets look at the gene usage data with ggplot2.

```{r, fig.height = 4, fig.width = 6}
# we can compute the total number of rearrangements per sample
total.usage <- aggregate(gene_usage_count~sample_id+condition,
                         FUN = sum, data = Ig)
total.usage$total <- total.usage$gene_usage_count
total.usage$gene_usage_count <- NULL

# merge it with the original data
viz <- merge(x = Ig, y = total.usage,
             by = c("sample_id", "condition"),
             all.x = TRUE)

# compute %
viz$gene_usage_pct <- viz$gene_usage_count/viz$total*100

# visualize
ggplot(data = viz)+
  geom_point(aes(x = gene_name, y = gene_usage_pct, fill = condition, 
                 shape = condition), stroke = 0, size = 3,
             position = position_jitterdodge(jitter.width = 0.25, 
                                             dodge.width = 0.75))+
  theme_bw(base_size = 11)+
  ylab(label = "Usage [%]")+
  xlab(label = '')+
  theme(legend.position = "top")+
  scale_fill_manual(name = "Condition", values = c("orange", "#6666ff"))+
  scale_shape_manual(name = "Condition", values = c(21, 22))
```


## Analysis with IgGeneUsage
Lets analyze the differential gene usage in the data Ig. We use the 
default values of most DGU arguments.

```{r}
M2 <- DGU(usage.data = Ig,
         mcmc.warmup = 250,
         mcmc.steps = 1000,
         mcmc.chains = 2,
         mcmc.cores = 1,
         hdi.level = 0.95,
         adapt.delta = 0.95,
         max.treedepth = 12)
```




## Model checking

### Checking MCMC sampling

  * divergences, tree-depth, energy

```{r}
rstan::check_hmc_diagnostics(M2$glm)
```

  * Rhat and Neff

```{r, fig.height = 3, fig.width = 6}
gridExtra::grid.arrange(rstan::stan_rhat(object = M2$glm),
                        rstan::stan_ess(object = M2$glm),
                        nrow = 1)
```


### Posterior predictive checks: repertoire-specific

```{r, fig.height = 5, fig.width = 8}
ggplot(data = M2$ppc$ppc.repertoire)+
  facet_wrap(facets = ~sample_name, ncol = 4)+
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", col = "darkgray")+
  geom_errorbar(aes(x = observed_count, y = ppc_mean_count, 
                    ymin = ppc_L_count, ymax = ppc_H_count), col = "darkgray")+
  geom_point(aes(x = observed_count, y = ppc_mean_count, 
                 fill = condition), shape = 21, size = 1)+
  theme_bw(base_size = 11)+
  theme(legend.position = "top")+
  scale_x_log10(breaks = c(1, 10, 100, 1000), 
                labels = expression(10^0, 10^1, 10^2, 10^3))+
  scale_y_log10(breaks = c(1, 10, 100, 1000), 
                labels = expression(10^0, 10^1, 10^2, 10^3))+
  xlab(label = "Observed usage [counts]")+
  ylab(label = "Predicted usage [counts]")+
  annotation_logticks(base = 10, sides = "bl")
```




### Posterior predictive checks: overall

```{r, fig.height = 4, fig.width = 6}
ggplot(data = M2$ppc.data$ppc.gene)+
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", col = "darkgray")+
  geom_errorbar(aes(x = observed_prop*100, ymin = ppc_L_prop*100, 
                    ymax = ppc_H_prop*100), col = "darkgray")+
  geom_point(aes(x = observed_prop*100, y = ppc_mean_prop*100, 
                 fill = condition), shape = 21, size = 1)+
  theme_bw(base_size = 11)+
  theme(legend.position = "top")+
  xlab(label = "Observed usage [%]")+
  ylab(label = "Predicted usage [%]")+
  scale_x_log10()+
  scale_y_log10()+
  annotation_logticks(base = 10, sides = "bl")
```




## Results
Main results:

```{r}
kable(x = M2$glm.summary, row.names = FALSE, digits = 3)
```

The effect size and $\pi$ are related. Lets visualize them for each gene (shown 
as a point). There are no genes with $\pi \approx 1$ $\rightarrow$ no sufficiently
strong evidence of differential gene usage based on IgGeneUsage. 

```{r, fig.height = 4, fig.width = 6}
# format data
stats <- M2$glm.summary
stats <- stats[order(abs(stats$es_mean), decreasing = FALSE), ]
stats$gene_fac <- factor(x = stats$gene_name, levels = stats$gene_name)

stats <- merge(x= stats, y = M2$test.summary, by = "gene_name")

ggplot(data = stats)+
  geom_hline(yintercept = 0, linetype = "dashed", col = "gray")+
  geom_errorbar(aes(x = pmax, y = es_mean, ymin = es_L, ymax = es_H), 
                col = "darkgray")+
  geom_point(aes(x = pmax, y = es_mean))+
  geom_text_repel(data = stats, aes(x = pmax, y = es_mean, label = gene_fac))+
  theme_bw(base_size = 11)+
  xlab(label = expression(pi))+
  xlim(0, 1)
```



## Comparison with the Welch's t-test (T-test)

```{r, fig.height = 6, fig.width = 8}
ggplot()+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), 
             linetype = "dashed", col = "darkgray")+
  geom_point(data = stats, col = "red", size = 2,
             aes(x = pmax, y = -log10(t.test.fdr.pvalue)))+
  geom_text_repel(data = stats, 
                  aes(x = pmax, y = -log10(t.test.fdr.pvalue), 
                      label = gene_name), size = 4, 
                  min.segment.length = 0.1)+
  xlim(0, 1)+
  ylab(label = "-log10 (P-value) from t-test [FDR corrected]")+
  xlab(label = expression(pi))+
  theme_bw(base_size = 11)+
  theme(legend.position = "top")+
  scale_color_discrete(name = '')
```


### Outlier: IGHV5

```{r, fig.height = 5, fig.width = 8}
promising.genes <- unique(M2$usage.data$gene_names)

ppc.gene <- M2$ppc.data$ppc.gene
ppc.gene <- ppc.gene[ppc.gene$gene_name %in% promising.genes, ]

ggplot()+
  geom_errorbar(data = ppc.gene,
                aes(x = gene_name, ymin = ppc_L_prop*100, 
                    ymax = ppc_H_prop*100, col = condition),
                position = position_dodge(width = .8), width = 0.75)+
  geom_point(data = viz[viz$gene_name %in% promising.genes, ],
             aes(x = gene_name, y = gene_usage_pct, col = condition),
             shape = 21, size = 1.5, fill = "black",
             position = position_jitterdodge(jitter.width = 0.15, 
                                             jitter.height = 0, 
                                             dodge.width = 0.8))+
  theme_bw(base_size = 11)+
  theme(legend.position = "top")
```


### Outliers IGHV5 (original data)

```{r, fig.height = 4, fig.width = 7}
promising.genes <- "IGHV5"

ggplot(data = viz[viz$gene_name %in% promising.genes, ])+
  facet_wrap(facets = ~gene_name, ncol = 1, scales = "free_y")+
  geom_point(aes(x = sample_id, y = gene_usage_count, fill = condition, 
                 size = total/10^3), shape = 21)+
  theme_bw(base_size = 11)+
  ylab(label = "Usage [count]")+
  xlab(label = '')+
  theme(legend.position = "top")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4))+
  scale_fill_manual(name = "Condition", values = c("orange", "#6666ff"))+
  scale_size_continuous(name = expression("N ("*10^3*")"), range = c(1, 5))+
  ylab(label = "Usage [count]")+
  xlab(label = '')+
  scale_y_log10()+
  annotation_logticks(base = 10, sides = "l")
```





## Comparison with the Wilcoxon signed-rank test (U-test)
No evidence of differential usage (at FDR level 0.05)

```{r, fig.height = 4.5, fig.width = 8}
ggplot()+
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01)), 
             linetype = "dashed", col = "darkgray")+
  geom_point(data = stats, col = "red", size = 2,
             aes(x = pmax, y = -log10(u.test.fdr.pvalue)))+
  geom_text_repel(data = stats, 
                  aes(x = pmax, y = -log10(u.test.fdr.pvalue), 
                      label = gene_name), size = 4, 
                  min.segment.length = 0.1)+
  xlim(0, 1)+
  ylab(label = "-log10 (P-value) from U-test [FDR corrected]")+
  xlab(label = expression(pi))+
  theme_bw(base_size = 11)+
  theme(legend.position = "top")+
  scale_color_discrete(name = '')
```

# Case Study III: differential species abundance in microbial communities

 * (in planning phase)
 
 
 
# Leave-one-out based cross validation

* (in planning phase)